Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-10-04 paper.tex
pass through the introducion -- @lenacoll you can now work on the Intro

In this paper, we consider the $\rnni$ space on the ranked phylogenetic trees with all taxa being of rank $0$.
An example of such a tree is depicted in Figure~\ref{fig:...}.
In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is $\rnniu$ (see the Section~\ref{sex:...} for accurate definitions).

@@

In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry of the $\rnni$ space and also consider several computational versions of the problems.
Specifically, in this paper we establish the exact radius and diameter of the space, [... -- Lena, please finish this including the approaximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]

\section{Definitions}

A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of taxa for $n \in \mathbb N$, and a rank function that maps $X$ to $0$, all internal nodes to $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.

In the literature ranked trees are often referred to as labelled history in the context of coalescent trees.
The following notion differs from the definition of ranks in that context, since we do not use negative integers as ranks.
A \emph{rank function} on $T$ maps the vertex set of $T$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The rank of a node $v$ in a ranked tree will be denoted by $\rank(v)$.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.

We will identify leaves with their labels and also call them \emph{taxa} (singular: \emph{taxon}) following the systematic biology tradition.
All nodes that are no leafs are called \emph{internal nodes} and an edge $(u,v)$ is an \emph{internal edge}, if both $u$ and $v$ are internal nodes.
Our definition of ranked tree leads to a natural characterisation of edge lengths within ranked trees:
integer $l$ denotes the \emph{length} of an edge $e=(u,v)$ if $|\rank(v) - \rank(u)| = l$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

Now we are ready to define the tree space which is the subject of study in this paper, the $\rnni$ graph.
Note that we will define the $\rnni$ graph on ultrametric trees where all leaves have the same rank $0$, and therefore refers to the $\rnniu$ graph as introduced in \autocite{Gavryushkin2018-ol}.
The vertex set of the \emph{$\rnni$ graph} is given by the set of all ranked trees on $n$ taxa.
We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in this graph by an edge:
A swap of the ranks of two internal nodes ($x$ and $y$) of subsequent ranks ($|\rank(x) - \rank(y)| = 1$) is called a \emph{rank move}.
We say that trees $T$ and $R$ are \emph{connected} by an \emph{$\nni$ move} move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
The above definition follows \autocite{Gavryushkin2018-ol} and is equivalent to the common definition of $\nni$ operations as given in \autocite{Steel2016-ye}.
Trees that are connected by either a rank move or an $\nni$ move are called \emph{$\rnni$ neighbours} or \emph{adjacent in} $\rnni$.
Note that thus defined $\rnni$ graph is a metric space with the standard shortest-path distance.

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{RNNI}
	\caption{Four trees connected by edges in the $\rnni$ graph. A rank move of $\mrca(\{1,2\})$ and $\mrca(\{3,4\})$ and all $\nni$ moves on the edge highlighted in red are illustrated}
	\label{fig:RNNI}
\end{figure}

\todo{label of taxa in figure shouldn't be numbers}

The following definitions are less standard and required for the results obtained in this paper.
A set $\mathbb S$ of trees is called \emph{convex} if for all pairs of trees $T,R \in \mathbb S$, there exists a shortest path from $T$ to $R$ such that every tree on $p$ is element of $\mathbb S$.

@@

This paper is an advance towards the general goal of understanding the geometry of the $\rnni$ space.
In this section we are investigating the geometry by analysing the structure of shortest paths in $\rnni$.
We will start by presenting an algorithm for computing paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for our studies.
Using this algorithm, we are able to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is convex in $\rnni$.
This distinguishes $\rnni$ from $\nni$ where this does not hold.
One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
Finally, we disprove the so-called Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}, and formulate an alternative version of it, the Cluster Theorem (Section~\ref{section:split_theorem}).

@@

\begin{theorem}
	The $\rnni$ space on $n$ taxa is isomorphic to the space of maximum chains of the partition lattice $\Pi_n$ where two maximum chains are connected by an edge if, and only if, they differ by exactly one partition.
\end{theorem}

>>>>

In this paper, we consider the $\rnni$ space on the ranked phylogenetic trees with all taxa being of rank $0$.
An example of such a tree is depicted in Figure~\ref{fig:...}.
In the terminology of \autocite{Gavryushkin2018-ol}, the space considered in this paper is the space of ranked ultrametric phylogenetic trees $\rnniu$.
We postpone accurate definitions until later in this section.

@@

In line with the research programme proposed in \autocite{Gavryushkin2018-ol}, we investigate the geometry and algorithmic complexity of the $\rnni$ space.
Specifically, in this paper we establish the exact radius and diameter of the space, [... -- Lena, please finish this including the approaximate algorithm that works exactly on small spaces (Lena: is it already in the text?)]

In the rest of introduction we formally introduce non-standard definitions necessary for this paper.

A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of taxa for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all elements of $X$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
The latter means that if $x$ and $y$ are two internal nodes of $T$ such that there exists a path from a taxon $z \in X$ to $y$ which passes through $x$ then $\rank(x) < \rank(y)$.
Ranked trees $(T_1, r_1)$ and $(T_2, r_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
Since all trees in this paper are ranked trees, we will abuse the notation and drop the ranking from the notation.
We will also refer to a ranked tree simply by tree.
For a tree $T$, we will use $\rank_T$ to refer to its ranking.

Our definition of a ranked tree implies a natural notion of the edge length.
Indeed, a tree edge $(u,v)$ has length $l$ if $|\rank(v) - \rank(u)| = l$.

Now we are ready to define the tree space which is the subject of study in this paper, the $\rnni$ graph.

The vertex set of the $\rnni$ graph is the set of all ranked trees on $n$ taxa.
We introduce two types of operation ($\rnni$ \emph{moves}) on trees (see Figure~\ref{fig:RNNI}) and say that two trees are adjacent in the $\rnni$ graph is they are connected by an operation of either type.
The first type of operation is called a \emph{rank move} and defined by swapping the ranks of two internal nodes that are not adjacent in the tree.
Formally, if $x$ and $y$ are such nodes of a tree $T$ that $|\rank_T(x) - \rank_T(y)| = 1$ and $(x, y)$ is not an edge in $T$ then the tree $R$ obtained from $T$ by only changing $\rank_R(x) = \rank_T(y)$ and $\rank_R(y) = \rank_T(x)$ is said to be obtained by a rank move.
The second type of operation is called an $\nni$ \emph{move} and defined in the usual way, that is, two trees $T$ and $R$ are said to be connected by an $\nni$ move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the trees obtained by shrinking $e$ and $f$ to internal nodes coincide.

\todo{Add a bit more space in all figures at the bottom to have at least a standard interval above he caption.}
\todo{In this figure -- rm arrows (think how to place the RNNI edges), possibly avoid using colour, and highlight where the rank move is made (hollow nodes?)}
\todo{Two of these trees are caterpillar trees.
Do you think it would be helpful to draw them as caterpillars to prepare readers for the rest of the paper?}
\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{RNNI}
	\caption{Two types of operation that define edges of the $\rnni$ graph.}
	\label{fig:RNNI}
\end{figure}

\todo{label of taxa in figure shouldn't be numbers\\
AG: Well -- if you decide to change this, go through all the definitions VERY carefully to make sure we update everywhere.}

@@

In this section we analyse shortest paths in $\rnni$.
We will start by presenting an algorithm for computing paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for understanding the geometry of $\rnni$.
Specifically, this algorithm will enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
\todo{The rest of this paragraph should be carefully checked and be in the introduction.
Generally, you might want to think a bit more about a better structure of this paper into chapters.}
This distinguishes $\rnni$ from $\nni$ where this does not hold.
One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
Finally, we disprove the so-called Strong Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}.
However, the Weak Split Theorem remains a conjecture.
Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:split_theorem}).

@@

\begin{theorem}
The $\rnni$ graph on $n$ taxa is isomorphic to the graph of maximal chains of the partition lattice $\Pi_n$ where two maximal chains are connected by an edge if and only if they differ by exactly one partition.
The corresponding metric spaces are isometric.
\end{theorem}
